Investigation of ant cuticle dataset using image texture analysis

### 2022-09-21 paper/main.tex
proof version

\def\typeofarticle{Research Article} 
\def\currentvolume{2} 
\def\currentissue{2}
\def\currentyear{2022} 
\def\currentmonth{} 
\def\ppages{xx--xx} 
\def\DOI{xxxx} 
\def\Received{xx xx 2022} 
\def\Revised{xx xx 2022}
\def\Accepted{xx xx 2022} 
\def\Published{xx 2022}

@@

\author{%
    Noah Gardner\affil{1},
    John Paul Hellenbrand\affil{2},
    Anthony Phan\affil{1},
    Haige Zhu\affil{1},
    Zhiling Long\affil{4},
    Min Wang\affil{3},
    Clint Penick\affil{2},
    and Chih-Cheng Hung\affil{1}\corrauth
}%

@@

\address{%
    \addr{\affilnum{1}}{Laboratory of Machine Vision and Security Research, %
        College of Computing and Software Engineering, Kennesaw State %
        University, Marietta GA, USA}%
    \addr{\affilnum{2}}{College of Science and Mathematics, Kennesaw State
        University, Kennesaw, GA, USA}%
    \addr{\affilnum{3}}{Department of Mathematics, Kennesaw State University,
        Kennesaw, GA, USA}%
    \addr{\affilnum{4}}{Department of Informatics and Mathematics,
        Mercer University, Atlanta, GA, USA}%
}
\corraddr{chung1@kennesaw.edu}

@@

Due to the extensive number of insect species, manual exploration of
insect-based information is difficult and often requires expert training.
Therefore, automated entomology has attracted biologists and computer
scientists, and is expected to be a major contribution to the future of
insect-based research \cite{martineau_survey_2017}. One of the most commonly
used data types for insect analysis is image data. To develop an image-based
system for insect analysis, we can utilize existing work in general
image analysis methods.

@@

Lim et al. \cite{lim_performance_2017} apply a CNN-based algorithm for insect
classification. Lim et al. \cite{lim_performance_2017} classify a subset of
insect species and families based on the classes available in the ImageNet
dataset. ImageNet is a widely used dataset of images labeled by experts with
millions of images and thousands of categories \cite{deng_imagenet_2009}. In the
ImageNet dataset, there are some categories that specify the class of the insect
on a species level, e.g. \textit{monarch butterfly} and \textit{ringlet
    butterfly} as well as some categories that specify the class of the insect on a
family level, e.g. \textit{ant}, \textit{fly}, and \textit{bee}
\cite{deng_imagenet_2009}. Lim et al. \cite{lim_performance_2017} use a modified
AlexNet architecture and experiment with different numbers of kernels and their
effect the performance of the model. Glick et al. \cite{glick_insect_2016}
employ a similar approach by classifying 277 insect classes from ImageNet using
a hierarchical convolutional neural network. The results from Lim et al.
\cite{lim_performance_2017} and Glick et al. \cite{glick_insect_2016} suggest
that a CNN is capable of differentiating between different hierarchical classes
of insects.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{assets/images/smooth_full_collage.png}
    \caption{Examples of smooth cuticle texture ant images in the dataset after
        center cropping, from AntWeb \cite{perrichot_antweb_2012}.}
    \label{fig:smooth-cuticle-texture}
\end{figure}

\begin{figure}
    \centering
    \includegraphics[width=0.7\textwidth]{assets/images/rough_full_collage.png}
    \caption{Examples of rough cuticle texture ant images in the dataset after
        center cropping, from AntWeb \cite{perrichot_antweb_2012}.}
    \label{fig:rough-cuticle-texture}
\end{figure}

@@

\begin{table}[h]
    \centering
    \caption{Experimental results of different algorithms}
    \input{assets/results.tex}
    \label{tab:results}
\end{table}

\FloatBarrier

@@

\begin{figure}[h]
    \centering
    \begin{subfigure}{.45\textwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/plots/resnet101_gt_tsne.png}
        % \caption{Ground truth labels}
    \end{subfigure}
    \begin{subfigure}{.45\textwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/plots/resnet101_pred_tsne.png}
        % \caption{Predicted labels}
    \end{subfigure}
    \caption{t-SNE visualization of the embeddings of the second to last layer
        of the randomly initialized ResNet101 model trained on dataset.}
    \label{fig:resnet101_tsne}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \begin{subfigure}{.45\textwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/plots/fresnet101_gt_tsne.png}
        % \caption{Ground truth labels}
    \end{subfigure}
    \begin{subfigure}{.45\textwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/plots/fresnet101_pred_tsne.png}
        % \caption{Predicted labels}
    \end{subfigure}
    \caption{t-SNE visualization of the embeddings of the second to last layer
        of the fine-tuned ResNet101 model trained on dataset.}
    \label{fig:fresnet101_tsne}
\end{figure}

@@

\newcommand{\segmentedsubwidth}{0.24\textwidth}
\begin{figure}
    \centering
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/r103.png}
        \caption*{Rough texture}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/r111.png}
        \caption*{Rough texture}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/s105.png}
        \caption*{Smooth texture}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/s107.png}
        \caption*{Smooth texture}
    \end{subfigure}

    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/15/r103.png}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/15/r111.png}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/15/s105.png}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/15/s107.png}
    \end{subfigure}

    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/19/r103.png}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/19/r111.png}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/19/s105.png}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{/home/ngardn10/projects/cuticulus/paper/assets/images/kviews/19/s107.png}
    \end{subfigure}
    \caption{Raw images are shown with corresponding segmented outputs from
        K-views algorithm. The second row shows the results of using size of 15
        $\times$ 15 patches. The third row shows the results of using size of 19
        $\times$ 19 patches. In the images, the color red represents the
        background class, green represents the smooth class, and the blue
        represent the rough class.}
    \label{fig:kviews_segmented}
\end{figure}

\FloatBarrier Some classified results from the K-views are shown in Figure
\ref{fig:kviews_segmented}. A training set of 550 patches for background, 455
patches for rough, and 493 patches for smooth are used with patch sizes of 15
$\times$ 15 and 19 $\times$ 19. Hence, the three classes are utilized in the
clustering. These patches are extracted from two rough and two smooth images
with sizes of 608 $\times$ 608. The K-views algorithm converges early although
we set the maximum number of iterations to be 1000. These experimental results
are preliminary due to the minimal number of source images used and future work
will explore a more expansive approach to the problem.

\FloatBarrier
\section{Conclusion}
We have shown in this work that a deep learning approach and classical image
texture analysis methods can be used to automatically categorize ants based on
their cuticle texture, therefore supporting research on the evaluation of the
cuticle micro sculpturing function in future work. Our categorization system is
novel in the field of automated insect identification due to the broad number of
species captured by it. A model that is pre-trained on a diverse image task such
as ResNet can be transferred to our domain of texture analysis\footnote{All code
    is publicly available on GitHub \url{https://github.com/ngngardner/cuticulus}}.
However, a deep learning algorithm created specifically for the domain of
texture analysis had an F1 score of 92\% in our experiments. In future work, we
will continue to explore texture classes present in the dataset which are not
captured by the binary class system. We will also investigate application of
texture-based deep maching learning in image texture analysis.

>>>>

\def\typeofarticle{Research article} 
\def\currentvolume{2} 
\def\currentissue{2}
\def\currentyear{2022} 
\def\currentmonth{} 
\def\ppages{xx--xx} 
\def\DOI{10.3934/aci.2022xxx} 
\def\Received{22 July 2022} 
\def\Revised{31 August 2022}
\def\Accepted{01 September 2022} 
\def\Published{xx 2022}

@@

\author{%
    Noah Gardner\affil{1},
    John Paul Hellenbrand\affil{2},
    Anthony Phan\affil{1},
    Haige Zhu\affil{1},
    Zhiling Long\affil{4}, \\
    Min Wang\affil{3},
    Clint Penick\affil{2},
    and Chih-Cheng Hung\affil{1,}\corrauth
}%

@@

\address{%
    \addr{\affilnum{1}}{Laboratory of Machine Vision and Security Research, %
        College of Computing and Software Engineering, Kennesaw State %
        University, Marietta GA, USA}%
    \addr{\affilnum{2}}{College of Science and Mathematics, Kennesaw State
        University, Kennesaw, GA, USA}%
    \addr{\affilnum{3}}{Department of Mathematics, Kennesaw State University,
        Kennesaw, GA, USA}%
    \addr{\affilnum{4}}{Department of Informatics and Mathematics,
        Mercer University, Atlanta, GA, USA}%
}
\corraddr{Email: chung1@kennesaw.edu.}

@@

Due to the extensive number of insect species, manual exploration of
insect-based information is difficult and often requires expert training.
Therefore, automated entomology has attracted biologists and computer
scientists, and is expected to be a major contribution to the future of
insect-based research~\cite{martineau_survey_2017}. One of the most commonly
used data types for insect analysis is image data. To develop an image-based
system for insect analysis, we can utilize existing work in general
image analysis methods.

@@

Lim et al. \cite{lim_performance_2017} apply a CNN-based algorithm for insect
classification. Lim et al. \cite{lim_performance_2017} classify a subset of
insect species and families based on the classes available in the ImageNet
dataset. ImageNet is a widely used dataset of images labeled by experts with
millions of images and thousands of categories~\cite{deng_imagenet_2009}. In the
ImageNet dataset, there are some categories that specify the class of the insect
on a species level, e.g. \textit{monarch butterfly} and \textit{ringlet
    butterfly} as well as some categories that specify the class of the insect on a
family level, e.g. \textit{ant}, \textit{fly}, and \textit{bee}
\cite{deng_imagenet_2009}. Lim et al. \cite{lim_performance_2017} use a modified
AlexNet architecture and experiment with different numbers of kernels and their
effect the performance of the model. Glick et al. \cite{glick_insect_2016}
employ a similar approach by classifying 277 insect classes from ImageNet using
a hierarchical convolutional neural network. The results from Lim et al.
\cite{lim_performance_2017} and Glick et al. \cite{glick_insect_2016} suggest
that a CNN is capable of differentiating between different hierarchical classes
of insects.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{assets/images/smooth_full_collage.png}
    \caption{Examples of smooth cuticle texture ant images in the dataset after
        center cropping, from AntWeb \cite{perrichot_antweb_2012}.}
    \label{fig:smooth-cuticle-texture}
\end{figure}

\begin{figure}[h]
    \centering
    \includegraphics[width=0.7\textwidth]{assets/images/rough_full_collage.png}
    \caption{Examples of rough cuticle texture ant images in the dataset after
        center cropping, from AntWeb \cite{perrichot_antweb_2012}.}
    \label{fig:rough-cuticle-texture}
\end{figure}

@@

\begin{table}[h]
    \centering
    \caption{Experimental results of different algorithms}
    \small
    \input{assets/results.tex}
    \label{tab:results}
\end{table}

%\FloatBarrier

@@

\begin{figure}[h!]
    \centering
    \begin{subfigure}{.45\textwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/plots/resnet101_gt_tsne.png}
        % \caption{Ground truth labels}
    \end{subfigure}
    \begin{subfigure}{.45\textwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/plots/resnet101_pred_tsne.png}
        % \caption{Predicted labels}
    \end{subfigure}
    \caption{t-SNE visualization of the embeddings of the second to last layer
        of the randomly initialized ResNet101 model trained on dataset.}
    \label{fig:resnet101_tsne}
\end{figure}

@@

\begin{figure}[h!]
    \centering
    \begin{subfigure}{.45\textwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/plots/fresnet101_gt_tsne.png}
        % \caption{Ground truth labels}
    \end{subfigure}
    \begin{subfigure}{.45\textwidth}
        \includegraphics[width=1\linewidth]{thesis_assets/plots/fresnet101_pred_tsne.png}
        % \caption{Predicted labels}
    \end{subfigure}
    \caption{t-SNE visualization of the embeddings of the second to last layer
        of the fine-tuned ResNet101 model trained on dataset.}
    \label{fig:fresnet101_tsne}
\end{figure}

@@

\newcommand{\segmentedsubwidth}{0.24\textwidth}
\begin{figure}
    \centering
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{figs/r103.png}
        \caption*{Rough texture}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{figs/r111.png}
        \caption*{Rough texture}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{figs/s105.png}
        \caption*{Smooth texture}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{figs/s107.png}
        \caption*{Smooth texture}
    \end{subfigure}

    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{figs/15/r103.png}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{figs/15/r111.png}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{figs/15/s105.png}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{figs/15/s107.png}
    \end{subfigure}

    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{figs/19/r103.png}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{figs/19/r111.png}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{figs/19/s105.png}
    \end{subfigure}
    \begin{subfigure}{\segmentedsubwidth}
        \includegraphics[width=1\linewidth]{figs/19/s107.png}
    \end{subfigure}
    \caption{Raw images are shown with corresponding segmented outputs from
        K-views algorithm. The second row shows the results of using size of 15
        $\times$ 15 patches. The third row shows the results of using size of 19
        $\times$ 19 patches. In the images, the color red represents the
        background class, green represents the smooth class, and the blue
        represent the rough class.}
    \label{fig:kviews_segmented}
\end{figure}

\FloatBarrier Some classified results from the K-views are shown in Figure
\ref{fig:kviews_segmented}. A training set of 550 patches for background, 455
patches for rough, and 493 patches for smooth are used with patch sizes of 15
$\times$ 15 and 19 $\times$ 19. Hence, the three classes are utilized in the
clustering. These patches are extracted from two rough and two smooth images
with sizes of 608 $\times$ 608. The K-views algorithm converges early although
we set the maximum number of iterations to be 1000. In order to classify an
image with K-views, the input image is split into patches. Then, each patch is
assigned to a cluster based on the distance to the cluster. If there are more
`smooth' patches, than `rough' patches, then the image will be classified as
rough. These experimental results are preliminary due to the minimal number of
source images used and future work will explore a more expansive approach to the
problem.

\section{Conclusions}
We have shown in this work that a deep learning approach and classical image
texture analysis methods can be used to automatically categorize ants based on
their cuticle texture, therefore supporting research on the evaluation of the
cuticle micro sculpturing function in future work. Our categorization system is
novel in the field of automated insect identification due to the broad number of
species captured by it. A model that is pre-trained on a diverse image task such
as ResNet can be transferred to our domain of texture analysis\footnote{All code
    is publicly available on GitHub \url{https://github.com/ngngardner/cuticulus}}.
However, a deep learning algorithm created specifically for the domain of
texture analysis had an F1 score of 92\% in our experiments. In future work, we
will continue to explore texture classes present in the dataset which are not
captured by the binary class system. We will also investigate application of
texture-based deep maching learning in image texture analysis.

\section*{Conflict of interest}
All authors declare no conflicts of interest in this paper.
